\textbf{COMP0086 Summative Assignment}

### 2022-11-09 submission.tex
update submission file and add additional plots for q3

$$P(\textbf{p})) = \prod_{d=1}^D P(p_d)$$

@@

$$\frac{\partial\mathcal{L}(\textbf{p})}{\partial p_d} = -\log (B(\alpha, \beta)) + (\alpha-1)\log p_d + (\beta-1)\log(1-p_d))$$

@@

Displaying the learned parameters:
\begin{figure}[h]
\centering
\begin{minipage}{0.5\textwidth}
  \centering
  \includegraphics[scale=0.5]{outputs/q1/q1d}
  \caption{ML parameters}
  \label{fig:1d}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
  \centering
  \includegraphics[scale=0.5]{outputs/q1/q1e}
  \caption{MAP parameters}
  \label{fig:1e}
\end{minipage}
\end{figure}

Comparing the equations:

    $$\hat{\textbf{p}}^{MLE} = \frac{1}{N}\sum_{n=1}^{N} \textbf{x}^{(n)}$$

    $$\hat{\textbf{p}}^{MAP}  =  \frac{\alpha - 1 +\sum_{n=1}^{N} \textbf{x}^n}{(N+\alpha+\beta-2)}$$

We see that when the number of data points increases, the MAP estimate approaches $\frac{1}{N}\sum_{n=1}^{N} \textbf{x}^{(n)}$, the MLE. This result makes sense because as our data set gets bigger, are less reliant on our prior.  Moreover, if in our data set of images, a specific pixel in all of these images are white or all black, the MLE for that pixel will be binary. This may not be be representative of our intuitions about images, where there should be some non-zero probability of a pixel being black or white. Thus, introducing an appropriate prior will ensure that the probability of that pixel will never be zero or one. Moreover, a prior can also help ensure numerical stability during calculations. The logarithm of zero is negative infinity, so having probability zero can be problematic for log-likelihoods calculations. A prior can ensure non-zero probabilities. In our case, by using a small Beta distribution prior on each pixel, our parameter values are biased be close to 0.5 and to never be at the extremities 0 and 1. We can see this in our MAP parameters figure where the range of our parameters is smaller than the range for the ML parameters. Also, the MLE has values of zero whereas the MAP does not. Interestingly, when $\alpha=1$ and $\beta=1$, $\hat{\textbf{p}}^{MLE} = \hat{\textbf{p}}^{MAP}$, intuitively this is when the prior is a uniform distribution and so there aren't any biases on the location of $\textbf{p}$ and we recover the MLE. On the other hand, a misspecified prior can be problematic, as the estimated parameters might be skewed by the prior and not properly represent the underlying data generating process and this can be worse than using the MLE.
%In this example, we're using images of different digits as part of our data set. If a particular digit occurs more frequently in the data set, we will see that digit when visualising the MLE or MAP (the digit zero in this case), they simply provides the "average" image pixels. With a set of images with more equal representation of each digit, the parameter visualisation could be an image overlaying the different digits, which is not representative of the "most likely" image in the data. In this case where our data is likely clustered, using a multimodal likelihood is probably a better idea than the unimodal likelihood we've used.

    We can also visualise $\hat{\textbf{p}}^{MAP}-\hat{\textbf{p}}^{MLE}$ to see that for likelihoods greater than 0.5 in the MLE, the MAP has a lower value and for likelihoods less than 0.5, the MAP has a higher value, confirming our intutions.

  \end{figure}
  \begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{outputs/q1/q1e-mle-vs-map}
  \caption{$\hat{\textbf{p}}^{MAP}-\hat{\textbf{p}}^{MLE}$}
  \label{fig:1e}
  \end{figure}

@@

\section*{Question 2}

@@

$$P(\textbf{x}^{(n)|\textbf{p}^{(1)}}=[0.5, 0.5,...,0.5]^T) = \prod_{d=1}^{D} (0.5)^{x_d^{(n)}} (0.5)^{1-x_d^{(n)}}$$

@@

$$P(\textbf{x}^{(n)}|\textbf{p}^{(2)}}=[p_d, p_d, ..., p_d]^T) = \prod_{d'=1}^{D} p_d^{x_{d'}^{(n)}} (1-p_d)^{1-x_{d'}^{(n)}}$$

@@

$$P(\textbf{x}^{(n)}|\textbf{p}^{(3)}}=[p_1, p_2, ..., p_D]^T) = \prod_{d=1}^{D} p_d^{x_d^{(n)}} (1-p_d)^{1-x_d^{(n)}}$$

For each model $M_i$, we can marginalise out $P$ to get $P( \{\textbf{x}^{(n)}\}_{n=1}^{N}|M_i)$:

@@

We only ave three models, so in this case the normalisation $P(D)$ can be expressed as a sum:

@@

\begin{center}
\begin{tabular}{l|c}%
 \bfseries $i$ & \bfseries$ P(M_i|\mathcal{D})$% specify table head
\csvreader[head to column names]{outputs/q2/q2c.csv}{}% use head of csv as column names
{\\\hline\csvcoli&\csvcolii}% specify your coloumns here
\end{tabular}
\end{center}

% TODO

@@

\item[(a)] The likelihood for a model consisting of a mixture of K multivariate Bernoulli distributions can be expressed as the product across $N$ data points:

$$P(\{\textbf{x}^{(n)}\}_{n=1}^{N}|\theta) = \prod_{i=1}^{N}P(x_i|\theta)$$

where $\{\textbf{x}^{(n)}\}_{n=1}^{N}$ is our data set with $\textbf{x}^{(n)} \in \mathbb{R}^{D \times 1}$ and $\theta = \{\pi, \mathbf{P}\}$, $\pi = [\pi_1, . . . , \pi_K] \in \mathbb{R}^{K \times 1}$$ our mixing proportions $(0 \leq \pi_k \leq 1; \sum_k \pi_k = 1)$ and $\mathbf{P} \in \mathbb{R}^{D \times K}$ the K Bernoulli parameter vectors with elements $p_{kd}$ denoting the probability that pixel d takes value 1 under mixture component k. We are also the images are iid under the model, and that the pixels are independent of each other within each component distribution.

For each $P(\textbf{x}^{(n)}|\theta)$:

$$P(\textbf{x}^{(n)}|\theta) = \sum_{k=1}^{K} \pi_k \prod_{d=1}^D (p_{kd})^{\textbf{x}^{(n)}_{d}} (1-p_{kd})^{1-\textbf{x}^{(n)}_{d}}$$

The log-likelihood $\mathcal{L}(\textbf{x}^{(n)}|\theta)$ can be expressed in matrix form:

$$\mathcal{L}(\textbf{x}^{(n)}|\theta) = \log \sum_{k=1}^{K}  \pi_k \exp\bigg(\textbf{x}^{(n)}\log(\mathbf{P}_{k}) + (1-\textbf{x}^{(n)})\log(1-\mathbf{P}_{k})\bigg) $$

which can be further vectorised using Python scipy's $logsumexp$ operation.

Moreover, the log-likelihood $\mathcal{L}(\{\textbf{x}^{(n)}\}_{n=1}^{N}|\theta)$ can be expressed:

$$\mathcal{L}(\{\textbf{x}^{(n)}\}_{n=1}^{N}|\theta) = \sum_{i=1}^{N} \Bigg( \log \sum_{k=1}^{K}  \pi_k \exp\bigg(\textbf{x}^{(n)}\log(\mathbf{P}_{k}) + (1-\textbf{x}^{(n)})\log(1-\mathbf{P}_{k})\bigg) \Bigg)$$

@@

We know that:

$$P(A|B) \propto P(B|A)P(A)$$

@@

$$P(s^{(n)} = k|x^{(n)}, \pi, \mathbf{P}) \propto  P(x^{(n)} | s^{(n)} = k, \pi, \mathbf{P}) P(s^{(n)} = k|\pi, \mathbf{P})$$

where $s^{(n)} \in \{1,...,K\}$ a discrete latent variable where $P(s^{(n)} = k|x^{(n)} | \pi) = \pi_k$. Note that $P(s^{(n)} = k|x^{(n)} | \pi) = P(s^{(n)} = k|x^{(n)} | \pi, \mathbf{P})$ as $s^{(n)$ isn't dependent on $\mathbf{P}$.

Let $P(s^{(n)} = k|x^{(n)}, \pi, \mathbf{P}) \propto P(s^{(n)}$ be the unnormalised responsibility $\tilde{r}_{nk}$. Using the mixture for component $k$ and the likelihood function of component $k$:

$$\tilde{r}_{nk} = \pi_k \prod_{d=1}^D (p_{kd})^{x^{(n)}_{d}} (1-p_{kd})^{1-x^{(n)}_{d}}$$

Normalising across the components:

$$r_{nk} = \frac{\tilde{r}_{nk}}{\sum_{j=1}^K\tilde{r}_{nj}}$$

and $r_{nk}$, we have calculated $P(s^{(n)} = k|x^{(n)}, \pi, \mathbf{P})$ for the E step.

Moreover,

$$ \log \tilde{r}_{nk} = \log \pi_k + \sum_{d=1}^D \bigg( x^{(n)}_{d}\log(p_{kd}) + (1-x^{(n)}_{d}) \log(1-\exp(\log(p_{kd}))) \bigg)$$

$$\log r_{nk} = \log\tilde{r}_{nk} - \log\sum_{j=1}^K\exp(\log\tilde{r}_{nj})$$

which can be vectorised using Python scipy's $logsumexp$ operation.

\item[(c)] We know that the expectation log joint can be expressed:

$$\Bigg\langle \sum_n \log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P})\Bigg\rangle_{q(\{s^{(n)}\})} = \sum_{n=1}^N q(s^{(n)}) \log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P})$$

Let this quantity be $E$.

Filling in the appropriate quantities:

$$ E = \sum_n r_n^T[\log(\pi)+\log(\mathbf{P})^T x^{(n)} + \log(1-\mathbf{P})^T(1-x^{(n)})]$$

where $q(s^{(n)}) = r_n^T$ and $\log P(x^{(n)}, s^{(n)}) = \log(\pi)+\log(\mathbf{P})^T x^{(n)} + \log(1-\mathbf{P})^T(1-x^{(n)})$. Noting that $P(x^{(n)}, s^{(n)}| \pi, \mathbf{P}) = P(x^{(n)}|s^{(n)}) \pi, \mathbf{P}) P(s^{(n)})| \pi, \mathbf{P}) = \tilde{r}_{n}$

To maximise with respect to $\pi$ and $P$ for the M step, we want to take the derivative, set to zero, and solve for $\hat\pi$ and $\hat P$.

For the $k^{th}$ element of $\pi$:

$$\frac{\partial E}{\partial \pi_k} = \sum_n r_{nk} \frac{1}{\pi_m}$$

$$\frac{\partial E}{\partial \pi_k} + \lambda = 0$$

where $\lambda$ is a Lagrange multiplier ensuring that the mixing proportions sum to unity.

$$\hat \pi_m = \frac{1}{N}\sum_{n}r_{nk}$$

For the $dk^{th}$ element of P:

$$\frac{\partial E}{\partial P_{dk}} = \sum_n r_{nk}\frac{\partial}{\partial P_{dk}}[x^{(n)}_{d}\log P_{dk}+(1-x^{(n)}_{d}) \log(1-P_{dk})]$$

Simplying:

$$\frac{\partial E}{\partial P_{dk}} = \sum_n r_{nk}(\frac{x^{(n)}_{d}}{ P_{dk}}-\frac{1-x^{(n)}_{d}}{1-P_{dk}})$$

$$\frac{\sum_n  x^{(n)}_{d}r_{nk}}{ P_{dk}}-\frac{\sum_n r_{nk}-\sum_n  x^{(n)}_{d}r_{nk}}{1-P_{dk}} = 0$$

Solving for $P_{dk}$:

$$P_{dk}\sum_n r_{nk}-P_{dk}\sum_n x^{(n)}_{d}r_{nk} = \sum_n  x^{(n)}_{d}r_{nk}-P_{dk}\sum_n  x^{(n)}_{d}r_{nk}$$

$$\hat{P}_{dk} = \frac{\sum_n x^{(n)}_{d}r_{nk}}{\sum_n r_{nk}}$$

@@

$$\arg \max_{\pi, P} \Bigg\langle \sum_n \log P(x^{(n)}, s^{(n)}| \pi, P)\Bigg\rangle_{q(\{s^{(n)}\})}$$

thus obtaining an iterative update for the parameters $\pi$ and $P$ in the M-step of EM.

\newpage
\item[(d)] The Python code for the EM algorithm:
\lstinputlisting[language=Python]{src/solutions/q3.py}
\newpage
Run your algorithm on the data set for values of K in {2, 3, 4, 7, 10}. Plot the log likelihood as a function of the iteration number, and display the parameters found. [30 marks]

\item[(e)] Run the algorithm a few times starting from randomly chosen initial conditions. Do you obtain the same solutions (up to permutation)? Does this depend on K? Show the learned probability vectors as images.

Comment on how well the algorithm works, whether it finds good clusters (look at the cluster means and responsibilities and try to interpret them), and how you might improve the model. [10 marks]

>>>>

$$P(\textbf{p}) = \prod_{d=1}^D P(p_d)$$

@@

@@

Displaying the learned parameters:
\begin{figure}[h]
\centering
\begin{minipage}{0.5\textwidth}
  \centering
  \includegraphics[scale=0.4]{outputs/q1/q1d}
  \caption{ML parameters}
  \label{fig:1d}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
  \centering
  \includegraphics[scale=0.4]{outputs/q1/q1e}
  \caption{MAP parameters}
  \label{fig:1e}
\end{minipage}
\end{figure}

Comparing the equations:
    $$\hat{\textbf{p}}^{MLE} = \frac{1}{N}\sum_{n=1}^{N} \textbf{x}^{(n)}$$
    and
    $$\hat{\textbf{p}}^{MAP}  =  \frac{\alpha - 1 +\sum_{n=1}^{N} \textbf{x}^n}{(N+\alpha+\beta-2)}$$

As the number of data points increases, the MAP estimate approaches $\frac{1}{N}\sum_{n=1}^{N} \textbf{x}^{(n)}$, the MLE. This makes sense because as our data set gets bigger, we are less reliant on our prior.
Moreover, if a specific pixel in all of the images of our data set are white or all black, the MLE for that pixel will be binary.
This may not be representative of our intuitions about image pixels, as there should be some non-zero probability of a pixel being black or white.
By introducing an appropriate prior we can ensure that the probability of that pixel will never be exactly zero or one.
In our case, with a Beta(3,3) prior on each pixel, our parameter values are biased to be closer to 0.5 and to never be at the extremities 0 and 1.
We can see this in Figure $\ref{fig:1e}$ where the range of our parameters is smaller than the range of Figure $\ref{fig:1d}$.
Figure $\ref{fig:1e-comparison}$ visualises $\hat{\textbf{p}}^{MAP}-\hat{\textbf{p}}^{MLE}$ and we can see that for likelihoods greater than 0.5 in the MLE, the MAP has a lower value and for likelihoods less than 0.5, the MAP has a higher value, confirming our intuitions.

  \begin{figure}[h]
  \centering
  \includegraphics[scale=0.4]{outputs/q1/q1e-mle-vs-map}
  \caption{$\hat{\textbf{p}}^{MAP}-\hat{\textbf{p}}^{MLE}$}
  \label{fig:1e-comparison}
  \end{figure}

Priors can also help ensure numerical stability during calculations.
The logarithm of zero is negative infinity, so having if the MLE is zero it can be problematic for log-likelihoods calculations whereas MAP can ensure non-zero probabilities.
We can see in Figure $\ref{fig:1d}$ and Figure $\ref{fig:1e}$ that the MLE has parameter values of zero whereas the MAP does not.
Interestingly, when $\alpha=1$ and $\beta=1$, $\hat{\textbf{p}}^{MLE} = \hat{\textbf{p}}^{MAP}$.
Intuitively this is when the prior is a uniform distribution and so there aren't any biases on the location of $\textbf{p}$ and we recover the MLE.

On the other hand, a mis-specified prior can be problematic, as the estimated parameters might be skewed by the prior and not properly represent the underlying data generating process, this can result in parameter estimates that are worse than using the MLE.
%In this example, we're using images of different digits as part of our data set. If a particular digit occurs more frequently in the data set, we will see that digit when visualising the MLE or MAP (the digit zero in this case), they simply provides the "average" image pixels. With a set of images with more equal representation of each digit, the parameter visualisation could be an image overlaying the different digits, which is not representative of the "most likely" image in the data. In this case where our data is likely clustered, using a multimodal likelihood is probably a better idea than the unimodal likelihood we've used.

@@

\newpage
\section*{Question 2}

@@

$$P(\textbf{x}^{(n)|\textbf{p}^{(1)}}=[0.5, 0.5,...,0.5]^T, M_1) = \prod_{d=1}^{D} (0.5)^{x_d^{(n)}} (0.5)^{1-x_d^{(n)}}$$

@@

$$P(\textbf{x}^{(n)}|\textbf{p}^{(2)}}=[p_d, p_d, ..., p_d]^T, M_2) = \prod_{d'=1}^{D} p_d^{x_{d'}^{(n)}} (1-p_d)^{1-x_{d'}^{(n)}}$$

@@

$$P(\textbf{x}^{(n)}|\textbf{p}^{(3)}}=[p_1, p_2, ..., p_D]^T, M_3) = \prod_{d=1}^{D} p_d^{x_d^{(n)}} (1-p_d)^{1-x_d^{(n)}}$$

For each model $M_i$, we can marginalise out $\textbf{p}^{(i)}}$ to get $P( \{\textbf{x}^{(n)}\}_{n=1}^{N}|M_i)$:

@@

We only have three models, so in this case the normalisation $P(\{\textbf{x}^{(n)}\}_{n=1}^{N})$ can be expressed as a sum:

@@

\begin{center}
\begin{tabular}{l|c}%
 \bfseries $i$ & \bfseries$ P(M_i|\{\textbf{x}^{(n)}\}_{n=1}^{N})$% specify table head
\csvreader[head to column names]{outputs/q2/q2c.csv}{}% use head of csv as column names
{\\\hline\csvcoli&\csvcolii}% specify your coloumns here
\end{tabular}
\end{center}

We can see that for models specified to have the same parameter value for all pixels is very unlikely under the given data set. This makes sense because it is specifying models where the image is essentially blank, which is not reflective of the data. Moreover, $M_1$ specifies a specific value of 0.5 for all the parameters whereas $M_2$ specifies any value for all the parameters as long as it's the same. So the model $M_1$ is a subset of the models specified in $M_2$ and we can see this reflected in our probabilities when $P(M_2|\{\textbf{x}^{(n)}\}_{n=1}^{N}) > P(M_1|\{\textbf{x}^{(n)}\}_{n=1}^{N})$.

@@

\item[(a)] The likelihood for a model consisting of a mixture of K multivariate Bernoulli distributions can be expressed as the product across $N$ data points:
$$P(\{\textbf{x}^{(n)}\}_{n=1}^{N}|\theta) = \prod_{i=1}^{N}P(x_i|\theta)$$
where $\{\textbf{x}^{(n)}\}_{n=1}^{N}$ is our data set with $\textbf{x}^{(n)} \in \mathbb{R}^{D \times 1}$ and $\theta = \{\pi, \mathbf{P}\}$, $\pi = [\pi_1, . . . , \pi_K] \in \mathbb{R}^{K \times 1}$ our mixing proportions $(0 \leq \pi_k \leq 1; \sum_k \pi_k = 1)$ and $\mathbf{P} \in \mathbb{R}^{D \times K}$ the K Bernoulli parameter vectors with elements $p_{kd}$ denoting the probability that pixel d takes value 1 under mixture component k. We are also the images are iid under the model, and that the pixels are independent of each other within each component distribution.

For each $P(\textbf{x}^{(n)}|\theta)$:
$$P(\textbf{x}^{(n)}|\theta) = \sum_{k=1}^{K} \pi_k \prod_{d=1}^D (p_{kd})^{\textbf{x}^{(n)}_{d}} (1-p_{kd})^{1-\textbf{x}^{(n)}_{d}}$$
The log-likelihood $\mathcal{L}(\textbf{x}^{(n)}|\theta)$ can be expressed in matrix form:
$$\mathcal{L}(\textbf{x}^{(n)}|\theta) = \log \sum_{k=1}^{K}  \pi_k \exp\bigg(\textbf{x}^{(n)}\log(\mathbf{P}_{k}) + (1-\textbf{x}^{(n)})\log(1-\mathbf{P}_{k})\bigg) $$
which can be further vectorised using Python scipy's $logsumexp$ operation.

Moreover, the log-likelihood $\mathcal{L}(\{\textbf{x}^{(n)}\}_{n=1}^{N}|\theta)$ can be expressed:
$$\mathcal{L}(\{\textbf{x}^{(n)}\}_{n=1}^{N}|\theta) = \sum_{i=1}^{N} \Bigg( \log \sum_{k=1}^{K}  \pi_k \exp\bigg(\textbf{x}^{(n)}\log(\mathbf{P}_{k}) + (1-\textbf{x}^{(n)})\log(1-\mathbf{P}_{k})\bigg) \Bigg)$$

@@

We know that:
$$P(A|B) \propto P(B|A)P(A)$$
Thus,
$$P(s^{(n)} = k|x^{(n)}, \pi, \mathbf{P}) \propto  P(x^{(n)} | s^{(n)} = k, \pi, \mathbf{P}) P(s^{(n)} = k|\pi, \mathbf{P})$$
where $s^{(n)} \in \{1,...,K\}$ a discrete latent variable where $P(s^{(n)} = k|x^{(n)} | \pi) = \pi_k$. Note that $P(s^{(n)} = k|x^{(n)} | \pi) = P(s^{(n)} = k|x^{(n)} | \pi, \mathbf{P})$ as $s^{(n)$ isn't dependent on $\mathbf{P}$.

Let $P(s^{(n)} = k|x^{(n)}, \pi, \mathbf{P}) \propto P(s^{(n)}$ be the unnormalised responsibility $\tilde{r}_{nk}$. Using the mixture for component $k$ and the likelihood function of component $k$:
$$\tilde{r}_{nk} = \pi_k \prod_{d=1}^D (p_{kd})^{x^{(n)}_{d}} (1-p_{kd})^{1-x^{(n)}_{d}}$$
Normalising across the components:
$$r_{nk} = \frac{\tilde{r}_{nk}}{\sum_{j=1}^K\tilde{r}_{nj}}$$
and $r_{nk}$, we have calculated $P(s^{(n)} = k|x^{(n)}, \pi, \mathbf{P})$ for the E step.

Moreover,
$$ \log \tilde{r}_{nk} = \log \pi_k + \sum_{d=1}^D \bigg( x^{(n)}_{d}\log(p_{kd}) + (1-x^{(n)}_{d}) \log(1-\exp(\log(p_{kd}))) \bigg)$$
and
$$\log r_{nk} = \log\tilde{r}_{nk} - \log\sum_{j=1}^K\exp(\log\tilde{r}_{nj})$$
which can be vectorised as $\log \textbf{r}_{n}$ using Python scipy's $logsumexp$ operation.

\item[(c)] We know that the expectation log joint can be expressed:
$$\Bigg\langle \sum_n \log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P})\Bigg\rangle_{q(\{s^{(n)}\})} = \sum_{n=1}^N q(s^{(n)}) \log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P})$$
Let this quantity be $E$. Each term of $E$ can be expressed:
$$q(s^{(n)}) = \textbf{r}_{n}$$
and
$$\log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P}) = \log[P(x^{(n)}| s^{(n)}, \pi, \mathbf{P})P(s^{(n)}| \pi, \mathbf{P})]$$
which is the vectorised version of $\log\tilde{r}_{nk}$ from part (b) so:
$$\log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P}) =\log(\pi)+\log(\mathbf{P})^T x^{(n)} + \log(1-\mathbf{P})^T(1-x^{(n)})$$
Combining:
$$ E = \sum_n \textbf{r}_{n}^T[\log(\pi)+\log(\mathbf{P})^T x^{(n)} + \log(1-\mathbf{P})^T(1-x^{(n)})]$$

To maximise with respect to $\pi$ and $\mathbf{P}$ for the M step, we want to take the derivative, set to zero, and solve for $\hat\pi$ and $\hat P$.

For the $k^{th}$ element of $\pi$:
$$\frac{\partial E}{\partial \pi_k} = \sum_n r_{nk} \frac{1}{\pi_k}$$

The second derivative:

$$\frac{\partial E}{(\partial \pi_k)^2} = \sum_n r_{nk} \frac{-1}{(\pi_k)^2}$$

is always negative because $r_{nk}\geq 0$, $\sum_{n}r_{nk}=1$, \pi_{k}\geq 0$, and $\sum_{n}\pi_{k}=1$, ensuring a maximum in the next step.

We can calculate the maximiser:
$$\frac{\partial E}{\partial \pi_k} + \lambda = 0$$
where $\lambda$ is a Lagrange multiplier ensuring that the mixing proportions sum to unity.

Thus,
$$\hat \pi_k = \frac{\sum_{n}r_{nk}}{N}$$

For the $dk^{th}$ element of $\mathbf{P}$:

$$\frac{\partial E}{\partial \mathbf{P}_{dk}} = \sum_n r_{nk}\frac{\partial}{\partial \mathbf{P}_{dk}}[x^{(n)}_{d}\log \mathbf{P}_{dk}+(1-x^{(n)}_{d}) \log(1-\mathbf{P}_{dk})]$$

$$\frac{\partial E}{\partial \mathbf{P}_{dk}} = \sum_n r_{nk}(\frac{x^{(n)}_{d}}{ \mathbf{P}_{dk}}-\frac{1-x^{(n)}_{d}}{1-\mathbf{P}_{dk}})$$

Similar to Question 1, we can see that taking the derivative, the term in the brackets will always be less than zero and with $r_{nk}\geq 0$ and $\sum_{n}r_{nk}=1$, the second derivative will always be negative. This ensures that we have a maximum in the next step.

$$\frac{\sum_n  x^{(n)}_{d}r_{nk}}{ \mathbf{P}_{dk}}-\frac{\sum_n r_{nk}-\sum_n  x^{(n)}_{d}r_{nk}}{1-\mathbf{P}_{dk}} = 0$$

Solving for $\hat{\mathbf{P}}_{dk}$:

$$\hat{\mathbf{P}}_{dk}\sum_n r_{nk}-\hat{\mathbf{P}}_{dk}\sum_n x^{(n)}_{d}r_{nk} = \sum_n  x^{(n)}_{d}r_{nk}-\hat{\mathbf{P}}_{dk}\sum_n  x^{(n)}_{d}r_{nk}$$

@@

$$\hat{\mathbf{P}}_{dk} = \frac{\sum_n x^{(n)}_{d}r_{nk}}{\sum_n r_{nk}}$$

@@

$$\arg \max_{\pi, \mathbf{P}} \Bigg\langle \sum_n \log P(x^{(n)}, s^{(n)}| \pi, \mathbf{P})\Bigg\rangle_{q(\{s^{(n)}\})}$$

thus obtaining an iterative update for the parameters $\pi$ and $\mathbf{P}$ in the M-step of EM.

For numerical stability, we can compute the maximisation step for the MAP of $\mathbf{P}$,$\hat{\mathbf{P}}_{dk}^{MAP}$:
$$\frac{\partial E'}{\partial \mathbf{P}_{dk}} = 0$$
where
$$E' = \sum_{n=1}^N q(s^{(n)}) \log P(\mathbf{P} | \pi, x^{(n)}, s^{(n)})$$
and
$$\log P(\mathbf{P} | \pi, x^{(n)}, s^{(n)}) = \log P( x^{(n)}, s^{(n)} | \pi, \mathbf{P} ) + \log P(\mathbf{P}) - \log P(x^{(n)}, s^{(n)} | \pi)$$
Assuming the same independent Beta prior on each pixel of each component:
$$\log P(\mathbf{P}) = \sum_{k=1}^K \sum_{d=1}^D -\log (B(\alpha, \beta)) + (\alpha-1)\log \mathbf{P}_{dk} + (\beta-1)\log(1-\mathbf{P}_{dk})$$
and
$$\frac{\partial \log P(\mathbf{P})}{\partial \mathbf{P}_{dk}} = \frac{(\alpha-1)}{\mathbf{P}_{dk}} -  \frac{(\beta-1)}{1-\mathbf{P}_{dk}}$$`
Thus, the derivative can be expressed as:
$$\frac{\partial E'}{\partial \mathbf{P}_{dk}} =  \sum_n \Bigg(r_{nk} \bigg(\frac{\partial \log P( x^{(n)}, s^{(n)} | \pi, \mathbf{P} )}{\partial \mathbf{P}_{dk}} + \frac{\partial \log P(\mathbf{P})}{\partial \mathbf{P}_{dk}} \bigg) \Bigg)$$
Substituting the appropriate expressions:
$$\frac{\partial E'}{\partial \mathbf{P}_{dk}} =  \sum_n \Bigg(r_{nk} \bigg(\frac{x^{(n)}_{d}}{ \mathbf{P}_{dk}}-\frac{1-x^{(n)}_{d}}{1-\mathbf{P}_{dk}} + \frac{(\alpha-1)}{\mathbf{P}_{dk}} -  \frac{(\beta-1)}{1-\mathbf{P}_{dk}} \bigg) \Bigg)$$
Simplifying:
$$\frac{\partial E'}{\partial \mathbf{P}_{dk}} =  \frac{(\alpha-1)+\sum_n r_{nk} x^{(n)}_{d}}{ \mathbf{P}_{dk}}-\frac{(\beta-1)+\sum_n r_{nk} (1-x^{(n)}_{d})}{1-\mathbf{P}_{dk}}$$

This form is very similar to Question 2 (c). By a similar argument, we can see that the second derivative will always be negative (additionally $r_{nk}\geq 0$ and $\sum_n r_{nk}=1$ ), ensuring a maximum in the calculation of the next step.

Setting $\frac{\partial E'}{\partial \mathbf{P}_{dk}}=0$ we can calculate $\hat{\mathbf{P}}_{dk}^{MAP}$:

$$\hat{\mathbf{P}}_{dk}^{MAP} = \frac{\alpha - 1 +\sum_n r_{nk} x^{(n)}_{d}}{(N+\alpha+\beta-2)}$$

\newpage
\item[(d)] Plotting the posterior likelihood as a function of the iteration number:

\begin{figure}[h]
\centering
\includegraphics[scale=0.35]{outputs/q3/q3e-0-log-pos}
\caption{Log Likelihood vs Iteration Number}
\label{fig:3d-log-like}
\end{figure}

where $epsilon$ is the stopping condition for the posterior posterior converges.

Displaying the parameters found for $K$ in $\{2, 3, 4, 7, 10\}$:
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{outputs/q3/q3e-0-initialised-p}
  \caption{Randomly initialised parameters}
  \label{fig:3d-initialised-p}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{outputs/q3/q3e-0-optimised-p}
  \caption{EM optimised parameters}
  \label{fig:3d-optimised-p}
\end{figure}

%\begin{minipage}{0.5\textwidth}

%\end{minipage}%
%\begin{minipage}{0.5\textwidth}
%  \centering
%  \includegraphics[scale=0.3]{outputs/q3/q3e-0-optimised-p}
%  \caption{EM optimised parameters}
%  \label{fig:3d-optimised-p}
%\end{minipage}
%\end{figure}

\newpage
The Python code for the EM algorithm:
\lstinputlisting[language=Python]{src/solutions/q3.py}
\newpage
\item[(e)] Running the algorithm a few times starting from randomly chosen initial conditions and visualising the parameters:

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{outputs/q3/q3e-0-optimised-p}
  \caption{EM optimised parameters: Trial 0}
  \label{fig:3e-initialised-p-trial-0}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{outputs/q3/q3e-1-optimised-p}
  \caption{EM optimised parameters: Trial 1}
  \label{fig:3e-initialised-p-trial-1}
\end{figure}
\newpage

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{outputs/q3/q3e-2-optimised-p}
  \caption{EM optimised parameters: Trial 2}
  \label{fig:3e-initialised-p-trial-2}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{outputs/q3/q3e-3-optimised-p}
  \caption{EM optimised parameters: Trial 3}
  \label{fig:3e-initialised-p-trial-3}
\end{figure}

%\begin{figure}[h]
%\centering
%\begin{minipage}{0.5\textwidth}
%  \centering
%  \includegraphics[scale=0.5]{outputs/q3/q3e-0-optimised-p}
%  \caption{EM optimised parameters: Trial 0}
%  \label{fig:3e-initialised-p-trial-0}
%\end{minipage}%
%\begin{minipage}{0.5\textwidth}
%  \centering
%  \includegraphics[scale=0.5]{outputs/q3/q3e-1-optimised-p}
%  \caption{EM optimised parameters: Trial 1}
%  \label{fig:3e-initialised-p-trial-1}
%\end{minipage}
%\end{figure}
%\begin{figure}[h]
%\centering
%\begin{minipage}{0.5\textwidth}
%  \centering
%  \includegraphics[scale=0.5]{outputs/q3/q3e-2-optimised-p}
%  \caption{EM optimised parameters: Trial 2}
%  \label{fig:3e-initialised-p-trial-2}
%\end{minipage}%
%\begin{minipage}{0.5\textwidth}
%  \centering
%  \includegraphics[scale=0.5]{outputs/q3/q3e-3-optimised-p}
%  \caption{EM optimised parameters: Trial 3}
%  \label{fig:3e-initialised-p-trial-3}
%\end{minipage}
%\end{figure}

For smaller k, we can visually see that we obtain very similar solutions (a 7 and a 0 for $k=2$). However for higher K, we see that this may not always be the case. For Trial 0 of $k=10$, we have two 0's whereas in Trial 1 we have three 0's. Interestingly, the zeros are all different, representing different variants of the written digit (i.e. a slanted zero, a slightly slanted zero, and a symmetric zero).

Moreover, looking at the responsibilities of each mixture component, we can see that when k is relatively small they are relatively evenly distributed. However for $k=7$ and especially $k=10$, we can see some components have very small or zero probability (i.e. $\pi_1$ and $\pi_2$ of trial 2). The corresponding parameter visualisations will essentially never be utilised because no probability is assigned to the component. This can be verified when we perform a TSNE visualisation of the responsibility vector for each of the images. We can see that for large k, qualitatively the number of clusters no longer matches the k value, indicating that some clusters are redundant. For example for $k=7$ and $k=10$ we can only qualitatively see four or five clusters with TSNE.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.35]{outputs/q3/q3e-0-tsne}
  \caption{TSNE Visualisation of Image responsibilities: Trial 0}
  \label{fig:3e-tsne-0}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.35]{outputs/q3/q3e-1-tsne}
  \caption{TSNE Visualisation of Image responsibilities: Trial 1}
  \label{fig:3e-tsne-1}
\end{figure}
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.35]{outputs/q3/q3e-2-tsne}
  \caption{TSNE Visualisation of Image responsibilities: Trial 2}
  \label{fig:3e-tsne-2}
\end{figure}
\newpage

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.35]{outputs/q3/q3e-3-tsne}
  \caption{TSNE Visualisation of Image responsibilities: Trial 3}
  \label{fig:3e-tsne-3}
\end{figure}
